### 2015-10-06 WOLFHPC15/article/introduction.tex

Many solutions to automatically optimize and parallelize numerical simulations have been proposed in the literature, either as specific frameworks~\cite{CPE:CPE3494,Trilinos-Overview,petsc-efficient} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation (such as fluid dynamics, magneto-hydrodynamics, molecular dynamics etc.) is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations~\cite{Ragan-Kelley:2013:HLC:2491956.2462176}. The DSL proposed in this paper, called \emph{Multi-Stencil Language} or MSL, is a DSL to generate a parallel component-based structure of a numerical simulation, that we call a \emph{multi-stencil program}. As discussed in the related work, this work is complementary to the optimization and parallelization of a single stencil kernel and is independent from implementation choices (as for example the form of the studied mesh).

>>>>

Many solutions to automatically optimize and parallelize numerical simulations have been proposed in the literature, either as specific frameworks~\cite{CPE:CPE3494,petsc-efficient} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation (such as fluid dynamics, magneto-hydrodynamics, molecular dynamics etc.) is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations~\cite{Ragan-Kelley:2013:HLC:2491956.2462176}. The DSL proposed in this paper, called \emph{Multi-Stencil Language} or MSL, is a DSL to generate a parallel component-based structure of a numerical simulation, that we call a \emph{multi-stencil program}. As discussed in the related work, this work is complementary to the optimization and parallelization of a single stencil kernel and is independent from implementation choices (as for example the form of the studied mesh).

### 2015-10-06 WOLFHPC15/article/mscac.tex

The solution proposed in this paper is based on both a domain specific language to describe a multi-stencil program, \ie a complete numerical simulation, and on a compiler which parses the language, applies transformations for automatic parallelization, and dumps the representation to a component-based runtime.

>>>>

The solution proposed in this paper is based on both a domain specific language to describe a multi-stencil program, \ie a complete numerical simulation, and on a compiler which parses the language, applies transformations for automatic parallelization, and dumps the representation to a component-based runtime, ready-to-fill with the user code.

### 2015-10-06 rapport/formalism.tex

%\rmq{Pas très clair la suite; motivation? dom == some sub elements of M ?}
For simulations, data is discretized based on the mesh; but for a given mesh, there are various ways of mapping the data depending on the mesh entities on which the data is mapped.
\begin{mydef}
\todo{JB: vraiment, ta def dit que pour chaque couple (noeud, arete), tu associes un element de $D_i$ ...}
$D_i$ is a set of elements of a mesh $\mathcal{M}=(V,E)$, constructed by a function $dom_i$ which defines a precise association between $V$ and $E$, $dom_i : V \times E \rightarrow D_i$.
\end{mydef}

\begin{mydef}
\todo{JB: def alternative}
An entity $d$ of a mesh $\mathcal{M}=(V,E)$ is a subset of the mesh elements: $d\subset (V\cup E)$.
A kind of mesh entities $D$ is a set of entities: $D\subset \mathcal{P}(V\cup E)$
\end{mydef}
For example, in a 2D Cartesian mesh a first kind of entities are the cells ($D_0$), they can be defined as all sets containing exactly four vertices and four edges connected as a cycle. Another kind of entities are the vertices ($D_1$) that can be defined as all singletons formed of a single vertex each.

@@

In a numerical simulation a set of data elements, or quantities, are applied onto the mesh and represent the set of values to compute, or to use, for computation.
\begin{mydef}
A quantity is a function $\delta_t: D_{\delta t} \mapsto V_{\delta t}$ which associates each entity of its definition domain $d\in D_{\delta t}$ to a value $v\in V_{\delta t}$ where $V_{\delta t}$ is the data type, typically $\mathbb{R}$ or $\mathbb{C}$.
The set of quantities applied on the mesh is denoted $\Delta$.
In the rest of this paper, the entities on which a quantity $\delta_t$ is mapped is denoted $dom(\delta_t)=D_{\delta t}$.
\end{mydef}
Another approach closer to that taken in applied mathematics would have been to define $\delta$ functions over $D \times T$; however the approach chosen in this paper where a distinct function is defined for each time-step better matches the implementation of numerical simulations where the outer loop iterates over time and where only a few time-steps of quantity values are stored.

@@

\begin{mydef}
A numerical expression $\text{exp}_{wt}: dom(w_t)\times R\mapsto V_{wt}$ is a function that computes the value of a quantity $w_t(d)$ (write quantity) for a given entity $d \in dom(w_t)=D$, using the set of its input data $R=\{\delta 1_{t-1}, \delta 2_{t-1}, ...\} \subset \Delta$ (read quantities). It is defined as $\text{exp}_{wt}(d,r) \rightarrow w_t(d)$.
\end{mydef}

@@

\begin{mydef}
A computation $c$ of a numerical simulation is defined as $c(R,w,\text{exp})$, where $R \subset \Delta, w \in \Delta$ and $\text{exp}$ a numerical expression. %$D$ is one of the subsets $D_i \subset \mathcal{M}$, such that $w : D \rightarrow V$.
\end{mydef}
It has to be noticed that at each time iteration, all the elements of a mesh are computed. However, it happens that computations of the mesh elements are splitted in different domains, as for example the computation of the physical border. In this case additional $D_i$ can be specified for the mesh $\mathcal{M}$.

\begin{mydef}
The set of $n$ ordered computations of a numerical simulation is denoted $\Gamma = [c_i]_{0 \leq i \leq n-1}$, such that $\forall c_i,c_j \in \Gamma$, if $i \leq j$, then $c_i$ is computed before $c_j$, and $c_j$ can be computed only when $c_i$ is finished.
\end{mydef}

\begin{mydef}
A \textit{multi-stencil program} is defined by the quadruplet $\mathcal{MSP}(T,\mathcal{M},\Delta,\Gamma)$.
\end{mydef}
%If the number of computations in $\Gamma$ is $card(\Gamma)=n$, such that $\bigcup_{i=0}^{n-1}c_i = \Gamma$, then $\bigcup_{i=0}^{m-1}R_i \cup w_i \subseteq \Delta$.

>>>>

%\rmq{Pas très clair la suite; motivation? dom == some sub elements of M ?}
For simulations, data is discretized based on the mesh; but for a given mesh, there are various ways of mapping the data depending on the mesh entities on which the data is applied.
% \begin{mydef}
% \todo{JB: vraiment, ta def dit que pour chaque couple (noeud, arete), tu associes un element de $D_i$ ...}
% $D_i$ is a set of elements of a mesh $\mathcal{M}=(V,E)$, constructed by a function $dom_i$ which defines a precise association between $V$ and $E$, $dom_i : V \times E \rightarrow D_i$.
% \end{mydef}

\begin{mydef}
%\todo{JB: def alternative}
An entity $e$ of a mesh $\mathcal{M}=(V,E)$ is a subset of the mesh elements: $e\subset (V\cup E)$.
A kind of mesh entities $E$ is a set of entities: $E\subset \mathcal{P}(V\cup E)$
\end{mydef}
For example, in a 2D Cartesian mesh a first kind of entities are the cells ($D_0$), they can be defined as all sets containing exactly four vertices and four edges connected as a cycle. Another kind of entities are the vertices ($D_1$) that can be defined as all singletons formed of a single vertex.

@@

In a numerical simulation a set of data elements, or quantities, are applied onto the mesh and represent the set of values to compute, or to use, for computation.
\begin{mydef}
A quantity is a function $\delta: E_{\delta} \mapsto V_{\delta}$ which associates each entity of its definition domain $e\in E_{\delta}$ to a value $v\in V_{\delta}$ where $V_{\delta}$ is the data type, typically $\mathbb{R}$ or $\mathbb{C}$.
The set of quantities applied on the mesh is denoted $\Delta$.
In the rest of this paper, the entities on which a quantity $\delta$ is mapped is denoted $entity(\delta)=E_{\delta}$.
\end{mydef}
Another approach closer to that taken in applied mathematics would have been to define $\delta$ functions over $E_{\delta} \times T$; however the approach chosen in this paper where a distinct function is defined for each time-step better matches the implementation of numerical simulations where the outer loop iterates over time and where only a few time-steps of quantity values are stored.

@@

\begin{mydef}
A numerical expression $\text{exp}_{w}: entity(w)\times R\mapsto V_{w}$ is a function that computes the value of a quantity $w(e)$ (written quantity) for a given entity $e \in entity(w)=E$, using the set of its input data $R=\{e_{1}, e_{2}, ...\} \subset \Delta$ (read quantities). It is defined as $\text{exp}_{w}(e,r) \rightarrow w(e)$.
\todo{JB: tu n'introduis pas la notion de stencil ici?}
\end{mydef}

\begin{mydef}
A computation kernel $k$ of a numerical simulation is defined as $k(R,w,\text{exp})$, where $R \subset \Delta, w \in \Delta$ and $\text{exp}$ a numerical expression. %$D$ is one of the subsets $D_i \subset \mathcal{M}$, such that $w : D \rightarrow V$.
\end{mydef}
% It has to be noticed that at each time iteration, all the values of a quantity are computed. However, it happens that computations of the quantity is split in different domains, as for example the computation of the physical border. In this case additional $D_i$ can be specified for the mesh $\mathcal{M}$.
% \todo{JB: je comprends pas vraiment...}

\begin{mydef}
The set of $n$ ordered computations of a numerical simulation is denoted $\Gamma = [k_i]_{0 \leq i \leq n-1}$, such that $\forall k_i,k_j \in \Gamma$, if $i \leq j$, then $k_i$ is computed before $k_j$, and $k_j$ can be computed when $k_i$ is finished.
%\todo{JB: je dirais que $c_j$ peut etre calculé quand $c_j$ l'est, j'eleverais le "only", c'est justement parce qu'il n'y a pas le only que tu peux faire du task parallelism}
\end{mydef}

@@

\begin{mydef}
A \textit{multi-stencil program} is defined by the sextuplet $\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma)$.
\end{mydef}
%If the number of computations in $\Gamma$ is $card(\Gamma)=n$, such that $\bigcup_{i=0}^{n-1}c_i = \Gamma$, then $\bigcup_{i=0}^{m-1}R_i \cup w_i \subseteq \Delta$.
